Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2016-10-02 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Completed Ch 5 and working Ch 6 edits

In the initial heat release rate experiments another heat flux gauge was used to measure the total heat flux at a location 2~m from the vertical centerline of the burners and 0.5~ above the top of the burners.  At this location the average peak total heat flux measured for the natural gas fires were approximately 0.42~kW/m$^2$.  The average peak values for the gasoline and polyurethane foam were approximately 0.74~kW/m$^2$ and 0.35~kW/m$^2$ respectively.

@@

Exposing the painted gypsum wallboard samples to the vertical radiant panel provided a wide range of physical damage to samples as shown in Figure~\ref{RPGB_SET}.  However based on visual observations there were no flaming ignitions.  For one of the samples at 15~kW/m$^2$, small sections of the painted paper surface on the gypsum wall board would lift or partially separate from the gypsum core and would begin to glow.  This process became more prevalent as the heat fluxes increased.  The radiant exposure of 25~kW/m$^2$ caused the samples to collapse as they were being removed from the holder.  Hence there were no photos of those samples.   Figure~\ref{RP_GB_Front} is a graph of the averaged time vs temperature curves for the three replicates at the 7 different heat flux levels.  Notice that the samples exposed to 17.5~kW/m$^2$ and higher have a temperature increase to a peak temperature followed by a decrease.  This trend may be indicative of the combustion of the paint and paper adding to the energy at the face of the sample. The temperatures at which the change in rate of temperature rise occur are between 300 and 400~ $^{\circ}$C.

@@

The cone calorimeter experiments generated heat release are per unit area values ranging from a low of approximately 160 kW/m$^2$ at an incident heat flux of 35 kW/m$^2$ to approximately 230 kW/m$^2$ at an incident heat flux of 75 kW/m$^2$.  The average total heat released during each burn was 2.8 kJ/m$^2$.

\begin{table}
	\centering
	\begin{tabular}{|c|c|}
		\hline Incident Heat Flux & Heat Release Rate per Unit Area    \\
		\hline (kW/m$^2$) & (kW/m$^2$)  \\ \hline
		\hline 35 	& 163 	\\
		\hline 50	& 179   \\
		\hline 70	& 255  	\\
		\hline 75	& 231 	\\
		\hline
	\end{tabular}
	\caption[Painted gypsum wallboard time heat release rate per unit area results]{Painted gypsum wallboard heat release rate per unit area results from the cone calorimeter. Each value is based on an average of three tests.}
	\label{tab:Gypsum wallboard_HRRA}
\end{table}

Seven replicate experiments were conducted in the radiant flooring panel apparatus.  The mean burn length was 432~mm with a 95$\%$ confidence limit of $\pm$18~mm. Comparing this burn length range, 414~mm to 450~mm, to the incident heat flux curve provided a critical heat flux of approximately 8.9~kW/m$^2$ with a 95$\%$ confidence limit of $\pm$~0.6~kW/m$^2$.  Photographs showing a radiant flooring panel test underway and sample nearing the flameout position are provided in Figure~\ref{FRP_Test_Photos}.

While the vertical radiant panel did not provide ignition data or heat release rate values it did provide information on the fire effect development on the painted gypsum wallboard samples with respect to heat flux.  Referring again to the five fire effect categories listed by DeHann~\cite{DeHaan:2012} and examining the Figure~\ref{RPGB_SET} we see surface thermal effects of discoloration on the painted paper surface of the gypsum wallboard samples at both the 10 and 12.5~kW/m$^2$ incident heat flux exposures.  The samples exposed to 15~kW/m$^2$ exhibit charring and given the thin nature of the painted paper surface penetration was evidenced on two of the three sample.  The sample of the right also shows signs of the painted paper being consumed.  The samples exposed to 17.5~kW/m$^2$ and above all show signs of significant levels of painted paper consumption.  As the heat flux was increased, the area of damage also increased. Even though the incident heat flux values were lower than the characteristic value measured at the center of the sample, as the heat flux in the center increased heat flux at the edges also increased above 15~kW/m$^2$. Hence the damaged area increased outward toward the edges. It would appear in this test series, that 15~kW/m$^2$ is right on the threshold of painted paper penetration and consumption.

DeHaan~\cite{DeHaan:2012} categorizes the fire effects that form patterns as follows:
\begin{enumerate}
	\item Surface deposits -– no irreversible effect on surface
	\item Surface thermal effects –- physical change such a discoloration or melting
	\item Charring -– evidence of surface burning
	\item Penetration -– charring below the surface
	\item Consumption –- loss of surface material, charring throughout
\end{enumerate}

Each apparatus provided similar but different thermal exposure conditions and each test method has a specific parameter or parameters that they were designed to measure for. Currently there is no specific protocol for a fire model user to follow to develop input data.  Nor is there a ``standard'' catalog of fire model inputs for a given material.  As a result the literature is sourced for information and users may need VII and VIII or to evaluate the validity of the calculations in those chapters.

@@

A photograph of the compartment fire experimental arrangement is shown in Figure~\ref{Compartment_Test}. In the instrumented gypsum wallboard experiments, as the burners were moved away from the position against the wall, the fire effects on the wall diminished significantly.  Therefore most of the fire pattern compartment experiments were conducted with the burners against the wall or corner of the compartment or less than one burner length from the wall.

Prior to each burn, grid lines approximately 51~mm on center were drawn on the painted gypsum wallboard panel that would be exposed to the fire.  After each burn, the fire patterns were photographed and analyzed for repeatability by looking at the portion of the grid that was burned. The fire damaged wallboard was subsequently removed, photographed again, and replaced with new sections of painted gypsum wallboard. Several types of fire pattern experiments were conducted in the compartments; to examine the impact of the wall construction on the resulting fire pattern; and to compare the fire patterns based on the burner centered on,or the rear wall, or in, or near a corner on the rear wall of the compartment.

@@

 The initial set of compartment experiments were conducted to examine if the type of wall construction would impact the development of the fire patterns. These experiments determined how the walls for the remaining compartment fire pattern experiments were constructed.  Three cases were examined: 1) Open back, 12.7~mm thick, regular core, painted gypsum board on the front of a wood frame only, 2) Closed back, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame representative of an interior wall, and 3) Insulated, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame, with insulation in the voids of the wood frame representative of an exterior wall. The insulation was 90~mm thick fiber glass batting,a kraft paper facing, and an ``R value'' of 13.  From the instrumented gypsum wallboard experiments, the peak temperatures on the backside of the wallboard occurred when the burners were positioned against the wall.  The peak temperatures were similar for all of the fuels, approximately 80~$^\circ$C, although the duration of the higher temperatures on the rear of the wallboard was longer for the natural gas and the gasoline fueled burners. Therefore the wall construction experiments were only conducted with the natural gas and gasoline fueled burners as the impact of insulating the back of the walls exposed to the fuels with the highest energy would produce the most significant differences, if any, to the fire patterns.

@@

Ten replicate experiments were conducted with each of the three wall construction types.  The walls were exposed to both the natural gas and the gasoline burners.  For these experiments the burner was in contact with the gypsum wallboard.  A total of 60 experiments were conducted.  The results will be presented by wall construction type.  The order of presentation will be open, closed, and insulated wall construction.

For this analysis, the outline of the fire pattern, or the line of demarcation~\cite{NFPA:921}, was defined as the points where charring, penetration and consumption of the paper covering the gypsum wallboard stopped. The grids were used to assist with the determination of the fire pattern area and the sizing of the images. This defining edge was used to measure the dimensions of the fire pattern, such as height and width. Given the irregular shape of the patterns, the area of each pattern could not be calculated based solely on the maximum height and maximum width measurements. Therefore the areas of the grid cells within the fire pattern were added up to determine the area of the fire patterns.  The analysis was carried out with a resolution of a quarter grid cell or approximately 0.003~m$^2$.

@@

Compartment experiments with the burner positioned in or near the corner were conducted with the open back wall construction.  The purpose of moving the burner position toward the corner was to examine the effect of reducing the available air entrainment area around the perimeter of the burner on the resulting fire pattern.  With the burner tight against both walls, the air entrainment perimeter was reduced by 50~\%. Previous studies on flame height with fires in corners have demonstrated the the flame height increased when compared to a similar fire positioned against a single wall or burning in the open~\cite{Poreh:1999,Saito:1993,Takahashi:1997,Kokkala:1993}.  The previous studies measured temperatures, heat flux, and mass flow rates for the purpose of developing correlations to predict flame height. The sole purpose of these experiments is to examine the fire patterns that result from well characterized fires on painted gypsum wall board.

@@

The compartment used for the corner position experiments was the same as the comapartment used for the wall positions experiments detailed in the previous section.  The burner was positioned in the left, rear corner against the face of two pieces of painted gypsum wallboard. As in the fire source characterization experiments the top of the fuel in each case was positioned 7.5~cm above the floor of the compartment. The source fires remained the same in terms of fuels, amounts of fuels used, burner size and burn duration.  The natural gas had a 300~s burn time which was approximately the upper bound for the total burn time for the 500~mL of gasoline.  By using the same burners, these compartment fire pattern experiments can be correlated with the instrumented wall experimental results.  Figure~\ref{GBNG34Sequence} displays a sequence of photographs taken during a natural gas burner corner experiment.  The photographs taken at a rate of five per second show the significant movement of the regions of visible combustion.  The grid drawn on the corner extends to 1.5 m above the top of the burner.

>>>>

In the initial heat release rate experiments another heat flux gauge was used to measure the total heat flux at a location 2~m from the vertical centerline of the burners and 0.5~m above the top of the burners.  At this location the average peak total heat flux measured for the natural gas fires were approximately 0.42~kW/m$^2$.  The average peak values for the gasoline and polyurethane foam were approximately 0.74~kW/m$^2$ and 0.35~kW/m$^2$ respectively.

@@

The cone calorimeter experiments generated peak heat release are per unit area values ranging from a low of approximately 160 kW/m$^2$ at an incident heat flux of 35 kW/m$^2$ to approximately 230 kW/m$^2$ at an incident heat flux of 75 kW/m$^2$.  The average total heat released during each burn was 2.8 MJ/m$^2$.  The total heat released during the cone calorimeter experiments compares favorably with the study by McGraw and Mowrer~\cite{McGraw:1999}, where the total heat released for the gypsum wallboard with 2 coats of latex paint was 2.3 to 2.7~MJ/m$^2$.

\begin{table}
	\centering
	\begin{tabular}{|c|c|}
		\hline Incident Heat Flux & Heat Release Rate per Unit Area    \\
		\hline (kW/m$^2$) & (kW/m$^2$)  \\ \hline
		\hline 35 	& 163 	\\
		\hline 50	& 179   \\
		\hline 70	& 255  	\\
		\hline 75	& 231 	\\
		\hline
	\end{tabular}
	\caption[Painted gypsum wallboard peak heat release rate per unit area results]{Painted gypsum wallboard peak heat release rate per unit area results from the cone calorimeter. Each value is based on an average of three tests.}
	\label{tab:Gypsum wallboard_HRRA}
\end{table}

Seven replicate experiments were conducted in the radiant flooring panel apparatus.  The mean burn length was 432~mm with a 95$\%$ confidence limit of $\pm$18~mm. Comparing this burn length range, 414~mm to 450~mm, to the incident heat flux curve provided a critical heat flux of approximately 8.9~kW/m$^2$ with a 95$\%$ confidence limit of $\pm$~0.6~kW/m$^2$.  Photographs showing a radiant flooring panel test underway and a close up of a sample near the location where the flame self extinguished are provided in Figure~\ref{FRP_Test_Photos}.

Exposing the painted gypsum wallboard samples to the vertical radiant panel provided a wide range of physical damage to samples as shown in Figure~\ref{RPGB_SET}.  However based on visual observations there were no flaming ignitions.  For one of the samples at 15~kW/m$^2$, small sections of the painted paper surface on the gypsum wall board would lift or partially separate from the gypsum core and would begin to glow.  This process became more prevalent as the heat fluxes increased.  The radiant exposure of 25~kW/m$^2$ caused all of the painted paper to be burned off of the sample and the exposure dehydrated the gypsum to the point where the samples collapsed as they were being removed from the apparatus.  Therefore no 25~kW/m$^2$ samples are shown in Figure~\ref{RPGB_SET}.   Figure~\ref{RP_GB_Front} is a graph of the averaged time vs temperature curves for the three replicates at the 7 different heat flux levels.  Notice that the samples exposed to 17.5~kW/m$^2$ and higher have noticable temperature increase to a peak temperature followed by a decrease.  This trend may be indicative of the combustion of the paint and paper adding to the energy at the face of the sample. The temperatures at which the change in rate of temperature rise occurred are between 300 and 400~$^{\circ}$C.  The temperature curve respresenting the 15~kW/m$^2$ experiments also demonstrates an increase in the rate of temperature rise beginning at approximately 300~$^{\circ}$C after more than 400~seconds of exposure.  Multiplying the time that each temperature peaked by the exposure heat flux yields the lowest total of heat was expended to create the most thermal damage, in the shortest time was at the 25~kW/m$^2$ exposure.  While at the 15~kW/m$^2$ exposure used more than twice the total amount of heat over a longer duration and yielded a lesser amount of damage.  The flux levels bounded by 15~kW/m$^2$ and 25~kW/m$^2$,have total energy exposures whihch are also bounded by the two extremes. This data indicates that depending on the exposure the amount of energy required to raise the temperature of the target material toward an ignition temperature, under these conditions, differs by a significant amount.

@@

While the vertical radiant panel did not provide ignition data or heat release rate values it did provide information on the fire effect development on the painted gypsum wallboard samples with respect to heat flux.  Referring again to the five fire effect categories listed by DeHann~\cite{DeHaan:2012} and examining the Figure~\ref{RPGB_SET} we see surface thermal effects of discoloration on the painted paper cover of the gypsum wallboard samples at both the 10 and 12.5~kW/m$^2$ incident heat flux exposures.  The samples exposed to 15~kW/m$^2$ exhibit charring and given the thin nature of the painted paper surface penetration was evidenced on two of the three sample.  The 15~kW/m$^2$ sample on the right also shows signs of the painted paper being consumed.  The samples exposed to 17.5~kW/m$^2$ and above all show signs of significant levels of painted paper consumption.  As the heat flux was increased, the area of damage also increased. Even though the incident heat flux values were lower than the characteristic value measured at the center of the sample, as the heat flux in the center increased heat flux at the edges also increased above 15~kW/m$^2$. Hence the damaged area increased outward toward the edges. It would appear in this test series, that 15~kW/m$^2$ heat flux exposure is on the threshold of painted paper penetration and consumption stage.

Each apparatus provided a thermal exposure to painted gypsum wallboard samples and an array of data was generated including: time to ignition, ignition temperature, heat release rate per unit area, and critical heat flux.  Each test method has a specific parameter or parameters that they were designed to measure for. As a result, the data presented in this chapter provides a wide range of values that can not be taken as a single number an applied.  In other words the data must be taken in context with how it was generated and analyzed accordingly. This is a challenge for fire investigators, analysts, and engineers alike.

The nature of the bench scale experimental data does not allow one to simply pick a value from a table an directly applying it as an input to a fire model.   Currently there is no specific protocol for a fire model user to follow to develop input data.  For example, there is no guidance on how to conduct a series of experiments to generate all of the data needed for the models.  Using the gypsum wall board as an example, material propeties such as density, specific heat, and thermal conductivity change over time as a result of the thermal exposure~\cite{Schroeder:1999,McGraw:1999}.  Of course this behavior is not unique to gypsum wallboard. The ideal case for the fire investigation user community would be a ``standard'' catalog of fire model inputs for a given material under a wide range of conditions.

There are other test methods that were not utilized here.  The lateral ignition and flame spread apparatus can be used to generate time to ignition and ignition temperature data in addition to critical heat flux for flame spread data.  The LIFT also considers time into the data where as the radiant flooring panel does not.  Further the LIFT and the cone calorimeter apparatus can be conducted in both horizontal and vertical orientation to better represent the end use of the sample material.  Thermogravimetric analysis (TGA) is another type of test which can be used to measure the mass loss rate of a material as a function of time and temperature.  Differential scanning calorimetry (DSC) is a test method which can be used to measure the specific heat capacity and in way phase change.  The TGA and DSC use samples that are on the order of a few grams.  This can make testing with a composite material as well as interpreting the results of a composite material difficult.

This chapter was anticipated to provide data that that correlated better than it does, although this should not have been a surprise given the wide range of thermal exposures, 10 to 75~kW/m$^2$.  The ignition temperatures measured in the cone calorimeter experiments conducted as part of this side do not align with the data in the published literature but the reason for the 200~$^{\circ}$C could not be resolved other than issues with thermocouple mounting, the use of paint, or changes to the gypsum board material properties over longer exposure times.

@@

A photograph of the compartment fire experimental arrangement is shown in Figure~\ref{Compartment_Test}. In the instrumented gypsum wallboard experiments, as the burners were moved away from the position against the wall, the fire effects on the wall diminished significantly.  Therefore the fire pattern compartment experiments were conducted with the burners against the wall or in the corner of the compartment.

Prior to each burn, grid lines approximately 51~mm on center were drawn on the painted gypsum wallboard panel that would be exposed to the fire.  After each burn, the fire patterns were photographed and analyzed for repeatability by looking at the portion of the grid that was burned. The fire damaged wallboard was subsequently removed, photographed again, and replaced with new sections of painted gypsum wallboard. Several types of fire pattern experiments were conducted in the compartments; to examine the impact of the wall construction on the resulting fire pattern; and to compare the fire patterns based on the burner centered on the rear wall, or in a corner on the rear wall of the compartment.

@@

 The initial set of compartment experiments were conducted to examine if the type of wall construction would impact the development of the fire patterns. These experiments determined how the walls for the remaining compartment fire pattern experiments were constructed.  Three cases were examined: 1) Open back, 12.7~mm thick, regular core, painted gypsum wallboard on the front of a wood frame only, 2) Closed back, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame representative of an interior wall, and 3) Insulated, 12.7~mm thick, regular core, painted gypsum wallboard on the front and back of a wood frame, with insulation in the voids of the wood frame representative of an exterior wall. The insulation was 90~mm thick fiber glass batting, with a kraft paper facing, and an ``R value'' of 13.  From the instrumented gypsum wallboard experiments, the peak temperatures on the backside of the wallboard occurred when the burners were positioned against the wall.  The peak temperatures were similar for all of the fuels, approximately 80~$^\circ$C, although the duration of the higher temperatures on the rear of the wallboard was longer for the natural gas and the gasoline fueled burners. Therefore the wall construction experiments were only conducted with the natural gas and gasoline fueled burners as the impact of insulating the back of the walls exposed to the fuels with the highest energy would produce the most significant differences, if any, to the fire patterns.

@@

Ten replicate experiments were conducted with each of the three wall construction types.  The walls were exposed to both the natural gas and the gasoline burners.  For all of these experiments the burner was in contact with the gypsum wallboard.  A total of 60 experiments were conducted.  The results will be presented by wall construction type.  The order of presentation will be open, closed, and insulated wall construction.

For this analysis, the outline of the fire pattern, or the line of demarcation~\cite{NFPA:921}, was defined as the points where charring, penetration and consumption of the paper covering the gypsum wallboard stopped. The grids were used to assist with the determination of the fire pattern area and the sizing of the images. This defining edge was used to measure the dimensions of the fire pattern, such as height and width. Given the irregular shape of the patterns, the area of each pattern could not be calculated based solely on the maximum height and maximum width measurements. Therefore the areas of the grid cells within the fire pattern were added up to determine the area of the fire patterns.  The analysis was carried out with a resolution of a quarter grid cell or approximately 0.001~m$^2$.

@@

Compartment experiments with the burner positioned in or near the corner were conducted with the open back wall construction.  The purpose of moving the burner position toward the corner was to examine the effect of reducing the available air entrainment area around the perimeter of the burner on the resulting fire pattern.  With the burner tight against both walls, the air entrainment perimeter was reduced by 50~\%. Previous studies on flame height with fires in corners have demonstrated the the flame height increased when compared to a similar fire positioned against a single wall or burning in the open~\cite{Poreh:1999,Saito:1993,Takahashi:1997,Kokkala:1993}.  The previous studies measured temperatures, heat flux, and mass flow rates for the purpose of developing correlations to predict flame height. The sole purpose of these experiments was to examine the fire patterns that result from well characterized fires on painted gypsum wall board.

@@

The compartment used for the corner position experiments was the same as the comapartment used for the wall positions experiments detailed in the previous section.  The burner was positioned in the left, rear corner against the face of two pieces of painted gypsum wallboard. As in the fire source characterization experiments the top of the fuel in each case was positioned 7.5~cm above the floor of the compartment. The source fires remained the same in terms of fuels, amounts of fuels used, burner size and burn duration.  The natural gas had a 300~s burn time which was approximately the upper bound for complete extinguishment (total burn time) for the 500~mL of gasoline and the polyurethane urethane block. Typically the gasiline and polyurethane burners were in decay approximately 120~seconds after ignition. By using the same burners, these compartment fire pattern experiments can be correlated with the instrumented wall experimental results.  Figure~\ref{GBNG34Sequence} displays a sequence of photographs taken during a natural gas burner corner experiment.  The photographs taken at a rate of five images per second show the significant movement of the regions of visible combustion.  The grid drawn on the corner extends to 1.5 m above the top of the burner.
